A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-20 rigid_bodies.tex
Introduction: check

Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations. This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators. The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two independent thermostat chains act separately on the rotational and the translational degrees of freedom. We have also simplified the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions (PBC). The expression we derive depends on the resultant interaction forces between body pairs. This avoids the evaluation of constraint forces, which are necessary if the pressure is calculated in terms of the interatomic forces.

This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion of rigid bodies using a Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we devise numerical integrators for both the NVE and NVT ensembles, whose correctness we discuss in terms of conserved quantities. Additionally, for the NVE case we present some results related to kinetic energy partition and, for the NVT case, we also analyze the statistical distributions of the sampled energy. After that, in Sec.~\ref{sec:pressure} we present the equation developed for the pressure.

Note that, throughout this paper, we employ the word ``quaternion'' as a synonym for a four-dimensional vector and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

@@

\section{Numerical Results}

@@

\section{Conclusion}

@@

If $\lVert \vt q \lVert^2 = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$. This is not true for a general operation in the opposite direction, $\vt x = \tr{\mt B} \vt y$. Besides unit norm, it is necessary that $\tr{\vt q}\vt y = 0$ (i.e. orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$. In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999}

As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
\[
{\mt B}_1 = \left[ \begin{smallmatrix}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallmatrix} \right], \,
{\mt B}_2 = \left[ \begin{smallmatrix}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallmatrix} \right], \text{ and }
{\mt B}_3 = \left[ \begin{smallmatrix}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallmatrix} \right].
\]

>>>>

Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and the length scales attainable in their atomistic simulations. This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so on. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators. The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and another one that corrects a pressure calculated in terms of interactions among free atoms.

The paper is organized as follows. In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT ensembles, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations. Thus, only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

@@

\section{Numerical Results}
\label{sec:numerical_results}

@@

\section{Concluding Remarks}
\label{sec:conclusion}

@@

If $\lVert \vt q \lVert^2 = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$. This is not true for a general operation in the opposite direction, $\vt x = \tr{\mt B} \vt y$. Besides unit norm, it is necessary that $\tr{\vt q}\vt y = 0$ (i.e. orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$. In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999} As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
\[
{\mt B}_1 = \left[ \begin{smallmatrix}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallmatrix} \right], \,
{\mt B}_2 = \left[ \begin{smallmatrix}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallmatrix} \right], \text{ and }
{\mt B}_3 = \left[ \begin{smallmatrix}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallmatrix} \right].
\]
